---
title: "Tutorial Project 5"
output: html_document
author: Darby Kleinert, Shannon McCollum, Julia Illiano
date: 21 Feb 2022
---
   This guided tutorial for Biol 202 will show the basics of how to use R and R studio to explore a data set, visualize the data, and perform basic statistical analysis of a hypothesis. Let's begin by learning about the main features of R studio that you will use. This window is where you can add and save any code chunks and text. To add code chunks, click on the green "C" located at the top of this window. To add comments, type anywhere in the white space within this box. 

Before we start exploring the data, we first need to load all the necessary packages. It is good practice to always load your packages first. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("UsingR")) install.packages("UsingR"); library(UsingR)
if (!require("cowplot")) install.packages("cowplot"); library(cowplot)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

##Data Exploration
   This tutorial will use a data set taken from a study published in the PLOS Journal that explored the effect of therapy dogs on the anxiety, depression, and pain rating of patients (Article link:https://doi.org/10.1371/journal.pone.0209232). 

We first need to load the data set into R studio using the read function.
```{r}
ERdogs <- read_csv("ERdogs.csv")
```
Now, we can look at the overall structure of the data set to see if all the variables loaded in properly:
```{r}
ERdogs #View the entire data set
head(ERdogs) #View the first ten columns
tail(ERdogs) #View the last ten columns 
str(ERdogs) #View each object 
```
Check to see if there are any NA values in your variables of interest. If there are, you can choose to keep them or remove them. In this case, we are removing NA values using this basic code:
```{r}
ERdogs <- drop_na(ERdogs, 'Patient  Anxiety rating based on image above')
ERdogs <- drop_na(ERdogs, 'Patient  Pain rating based on image above')
ERdogs <- drop_na(ERdogs, 'Patient  Depression rating based on image above')
ERdogs <- drop_na(ERdogs, `Heart Rate`)
```
Also, check to see that your variables of interest are showing up correctly. You will be working with categorical variables, continuous numeric variables (takes on any value), and discrete numeric variables (only takes on specific values). Make sure that categorical variables are showing up as characters, continuous numeric variables are showing up as numeric, and discrete numeric variables are showing up as factors. If they are not, use this format to change them:
```{r}
#Convert numeric to factor
ERdogs$`Patient  Anxiety rating based on image above`<- as.factor(ERdogs$`Patient  Anxiety rating based on image above`)
ERdogs$`Patient  Depression rating based on image above`<- as.factor(ERdogs$`Patient  Depression rating based on image above`)
ERdogs$`Patient  Pain rating based on image above`<- as.factor(ERdogs$`Patient  Pain rating based on image above`)
#Check to see that it converted 
str(ERdogs$`Patient  Anxiety rating based on image above`)
str(ERdogs$`Patient  Depression rating based on image above`)
str(ERdogs$`Patient  Pain rating based on image above`)
#Convert characters into numeric objects
ERdogs$`Heart Rate`<- as.numeric(ERdogs$`Heart Rate`)
#Check to see that it converted
str(ERdogs$`Heart Rate`)
```
If you are working with a larger data set, like this one, you can choose to select only certain variables to work with using the dplyr function. The "<-" symbol is called the assign operator and will be used to assign value and create new objects. In this case, we are making a subset of the data that contains only the selected variables. 
```{r}
ERdogs.subset <- dplyr::select(ERdogs, `Record ID`, Age, Gender, Group, `Patient  Anxiety rating based on image above`, `Patient  Pain rating based on image above`, `Patient  Depression rating based on image above`,`Heart Rate`)
head(ERdogs.subset)
tail(ERdogs.subset)
str(ERdogs.subset)
```

##Data Visualization
   Visual representations of data are useful to illustrate general trends and relationships within the data set. Choosing which graph to make will depend on how many variables you are exploring and whether your variables are categorical or numerical (continuous or discrete). 
   
A bar graph provides the best representation of a single categorical variable or a single discrete numerical variable. 
```{r}
ggplot(ERdogs)+
   aes(x = `Patient  Anxiety rating based on image above`) +
geom_bar() +
  theme_cowplot()+
  xlab("Anxiety Rating") + 
  ylab("Number of Participants")

ggplot(ERdogs) +
  aes(x = `Patient  Depression rating based on image above`) +
  geom_bar() +
  theme_cowplot()+
  xlab("Depression Rating") + 
  ylab("Number of Participants")

ggplot(ERdogs) +
  aes(x = `Patient  Pain rating based on image above`) +
  geom_bar() +
  theme_cowplot()+
  xlab("Pain Rating") + 
  ylab("Number of Participants")

#xlab labels the x-axis
#ylab labels the y-axis
```

A grouped bar graph is the best representation when comparing two categorical variables.
```{r}

ggplot(ERdogs) +
  aes(x = `Patient  Pain rating based on image above`, color = Group, fill = Group) +
  geom_bar(binwidth = 0.5, position = "dodge") +
  theme_cowplot()

ggplot(ERdogs) +
  aes(x = `Patient  Anxiety rating based on image above`, color = Group, fill = Group) +
  geom_bar(binwidth = 0.5, position = "dodge") +
  theme_cowplot()

ggplot(ERdogs) +
  aes(x = `Patient  Depression rating based on image above`, color = Group, fill = Group) +
  geom_bar(binwidth = 0.5, position = "dodge") +
  theme_cowplot()
```

A Histogram provides the best representation of a single continuous numerical variable. 
```{r}
ggplot(ERdogs, aes(x=`Heart Rate`)) + 
  geom_histogram(color="black", fill="white") +
  theme_cowplot()+
  xlab("Heart Rate (BPM)")
```


A strip chart or box plot would be best to use when comparing a numerical variable with a categorical variable. 
```{r}
# Strip Chart
ggplot(ERdogs, aes(x = Group, y= `Heart Rate`, fill = Group)) +
  geom_jitter()+
   xlab("Group") + 
  ylab("Heart Rate")

#Box plot
ggplot(ERdogs) +
  aes(x = Group , y = `Heart Rate` , fill=Group) +
  geom_boxplot() +
geom_jitter(color="black", size=0.4, alpha=0.9)+
  theme_cowplot()+
   xlab("Group") + 
  ylab("Heart Rate")

#geom_jitter shows all the data points on each graph
```

##Hypothesis Testing 
Shapiro testing checks to see if the data is normally distributed. You can only perform shapiro testing for numerical values not categorical.
```{r}
shapiro.test(ERdogs$`Heart Rate`)
shapiro.test(ERdogs$`Heart Rate` [ERdogs$`Patient  Pain rating based on image above`])
shapiro.test(ERdogs$`Heart Rate` [ERdogs$`Patient  Depression rating based on image above`])
shapiro.test(ERdogs$`Heart Rate` [ERdogs$`Patient  Anxiety rating based on image above`])
##all heart rates are normally distributed (p-val<0.05) when separated into each category

```
One way anova testing used to compare categorical(Pain/Anxiety/Depression Rating) and numerical (Heart Rate) variables. 
```{r}
one.way <- aov(`Heart Rate`~`Patient  Pain rating based on image above` , data=ERdogs)
summary(one.way)
one.way <- aov(`Heart Rate`~`Patient  Anxiety rating based on image above` , data=ERdogs)
summary(one.way)
one.way <- aov(`Heart Rate`~`Patient  Depression rating based on image above` , data=ERdogs)
summary(one.way)
##All p-values are greater than 0.05 which means there is no significant evidence between heart rate and different patient ratings. 

###OBSERVATIONS DELETED DUE TO MISSINGNESS???
```
Chi Square testing is used to see if two categorical variables have significant correlation. Here, we are seeing if the patient ratings have any correlation with which group they are in (dog or no dog.)
```{r}
chisq.test(ERdogs.subset$`Patient  Depression rating based on image above`, ERdogs.subset$Group)
chisq.test(ERdogs.subset$`Patient  Anxiety rating based on image above`, ERdogs.subset$Group)
chisq.test(ERdogs.subset$`Patient  Pain rating based on image above`, ERdogs.subset$Group)
##All p-values are greater than 0.05 which means there is no significant correlation between the variables.

#CHI SQUARE APPROXIMATION MAY BE INCORRECT???
```

##Assignment

After you thoroughly go through the three sections above, test your own knowledge by going through these steps: 

1) Choose new variables from the dataset and explore them using the tools given in Section I: Data Exploration. Pay close attention to the type of variables you are working with. 

2) Make a hypothesis using those new variables

3) To test the hypothesis, select appropriate graphs from Section 2: Data Visualization to visualize the data and make conclusions

4) Use the tools given in Section 3: Hypothesis Testing to use statistical tests to analyse your hypothesis and make final conclusions. Was your hypothesis supported? How can you tell? 


#Acknowledgements
https://data-flair.training/blogs/chi-square-test-in-r/
https://doi.org/10.1371/journal.pone.0209232
